---
title: "County-level covariate perparation"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages
```{r, message=FALSE}
library(tidyverse)
library(sf)
library(lubridate)
library(MetBrewer)
```

Load the data
```{r}
covar_grid <- st_read("geo-files/covar-grid.shp")

sc_counties <- st_read("geo-files/south-carolina-county-boundaries.shp") |> 
    st_transform(st_crs(covar_grid)) |> 
    rename(county=COUNTYNM)

shelters <- read_csv("data-proc/shelters-tidy-20-21.csv")

parks <- st_read("geo-files/parks-with-covars.shp") |> 
    rename(
        life_stage=lif_stg, land_cover=lnd_cvr, tree_canopy=tr_cnpy, min_temp=min_tmp, max_tmp=max_tmp,
        precipitation=prcpttn, jan_min_temp=jn_mn_t
    )
```

Aggregate the counts from shelters and parks to their matching county. We go ahead and complete the `life_stage` even though this is probably not needed, since occupancy status is defined at the species level. It may be useful if we end up using counts for some reason.

```{r}
occ_status <- function(life_stage, count) {
    if (all(is.na(count)))
        return(NA)
    
    found <- tibble(life_stage, count) |> 
        filter(!is.na(count) & count > 0)
    
    stages <- length(unique(found$life_stage))
    totals <- sum(found$count)
    
    if (stages > 1 | totals >= 6)
        return("established")
    if (totals > 0)
        return("suspected")
    return("absent")
    # found <- tibble(life_stage, count_shelters, count_parks) |> 
    #     filter(count_shelters > 0 | count_parks > 0)
    # 
    # stages <- length(unique(found$life_stage))
    # tot_shelters <- sum(found$count_shelters)
    # tot_parks <- sum(found$count_parks)
    # both_enc <- tot_shelters > 0 & tot_parks > 0
    # 
    # if (both_enc | stages > 1 | (tot_shelters + tot_parks) >= 6)
    #     return("established")
    # if ((tot_shelters + tot_parks) > 0)
    #     return("suspected")
    # return("absent")
}

shelters_agg <- left_join(shelters, sc_counties, by="county") |> 
    filter(genus %in% c("Amblyomma", "Ixodes", "Dermacentor") & species != "unknown") |> 
    select(county:count) |> 
    group_by(
        county, life_stage,
        tick_class=ifelse(genus == "Ixodes", str_c(genus, " spp."), str_c(genus, " ", species))
    ) |> 
    summarize(count=sum(count), .groups="drop") |> 
    # if a county pres., that means we got data from it, so make sure 0 ct appears for each group
    complete(county, tick_class, life_stage, fill=list(count=0))

parks_agg <- parks |> 
    st_join(sc_counties, st_within) |>
    st_drop_geometry() |> 
    group_by(
        county, life_stage,
        tick_class=ifelse(genus == "Ixodes", str_c(genus, " spp."), str_c(genus, " ", species))
    ) |> 
    summarize(count=sum(count), .groups="drop") |> 
    complete(county, tick_class, life_stage, fill=list(count=0))
    
county_occ <- parks_agg |> 
    full_join(shelters_agg, by=c("county", "tick_class", "life_stage"), suffix=c("_parks", "_shelters")) |> 
    group_by(county, tick_class) |> 
    mutate(
        occ_parks=occ_status(life_stage, count_parks), 
        occ_shelters=occ_status(life_stage, count_shelters),
        occ=case_when(
            occ_parks == "established" | occ_shelters == "established" ~ "established",
            occ_parks == "suspected" | occ_shelters == "suspected" ~ "suspected",
            TRUE ~ "absent"
        )
    ) |> 
    ungroup()
```

Join with the county geometries:
```{r}
county_occ <- left_join(select(sc_counties, county), county_occ) |> 
    # select(county, tick_class, ) |> 
    mutate(tick_class=map(tick_class, ~if(is.na(.x)) unique(county_occ$tick_class) else .x)) |> 
    unnest(tick_class)
```

Plot the result

```{r}
pal <- met.brewer("Kandinsky", type="discrete")

county_occ |> 
    rename(both=occ, `parks only`=occ_parks, `shelters only`=occ_shelters) |> 
    pivot_longer(`parks only`:both, "source") |> 
    mutate(value=fct_relevel(value, "established", after=Inf)) |>
    ggplot() +
    geom_sf(aes(fill=value), col="white") +
    facet_grid(source~tick_class) +
    scale_fill_manual(values=c(pal[4], pal[1], pal[2]), na.value=pal[3]) +
    labs(fill="Occupancy status") +
    theme_bw()

ggsave("figs/county-occupancy.pdf", width=8.75, height=6)
```


```{r}
covar_grid <- rename(
        covar_grid,
        land_cover=lnd_cvr, tree_canopy=tr_cnpy, min_temp=min_tmp, max_temp=max_tmp,
        precipitation=prcpttn, jan_min_temp=jn_mn_t
    )

covar_2020 <- filter(covar_grid, year == 2020)
```

```{r}
covar_grid_summ <- covar_grid |> 
    group_by(geo=as.character(geometry)) |> 
    summarise(across(c(land_cover:min_temp, max_temp:jan_min_temp), mean))

covar_grid_summ <- select(covar_grid_summ, -geo)
```

```{r}
agg_var <- function(var) {
    v <- pull(covar_grid_summ, {{var}})
    map_dbl(cont, ~mean(v[.x]))
}

perc_land_cover <- function(lcov) {
    map_dfr(cont, ~{
        lc <- factor(lcov[.x], levels=lc_levels)
        props <- table(lc) |> proportions() |> as.list()
        names(props) <- str_c("land_cover_", lc_levels)
        tibble_row(!!!props)
    })
}

cont <- st_contains(sc_counties, covar_grid_summ)
lc_levels <- unique(covar_grid_summ$land_cover)

county_covars <- sc_counties |> 
    transmute(
        county,
        !!!perc_land_cover(covar_grid_summ$land_cover),
        tree_canopy=agg_var(tree_canopy),
        min_temp=agg_var(min_temp),
        max_temp=agg_var(max_temp),
        precipitation=agg_var(precipitation),
        mean_rh=agg_var(mean_rh),
        jan_min_temp=agg_var(jan_min_temp)
    )
```

Since `precipitation` provided by PRISM is the cumulative precipitation over a month, note that this variable is the _mean_ monthly cumulative precipitation, over 2020 and 2021.

```{r}
ggplot(county_covars) +
    geom_sf(aes(fill=min_temp))
```

Add the covariates to `county_occ`
```{r}
county_occ <- left_join(county_occ, st_drop_geometry(county_covars), by="county")
```

Now we can save this for later.

```{r}
st_write(county_occ, "geo-files/county-model-data.shp", delete_dsn=TRUE)
```

