---
title: "Summary Statistics"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages and data
```{r}
library(tidyverse)
library(kableExtra)
library(ggcorrplot)
library(lubridate)
library(sf)
library(MetBrewer)

parks <- st_read("geo-files/parks-with-covars.shp") |> 
    rename(
        life_stage=lif_stg, land_cover=lnd_cvr, tree_canopy=tr_cnpy, min_temp=min_tmp, max_tmp=max_tmp,
        precipitation=prcpttn, jan_min_temp=jn_mn_t
    ) 
```

### Table: number of visits in which each species X life_stage combination were found

```{r}
parks |>
    st_drop_geometry() |> 
    filter(count > 0) |> 
    count(genus, species, life_stage, sort=TRUE)
```

### Figure: occupancy status of each species in surveyed parks

```{r}
occ_status_space <- function(date, life_stage, count) {
    vis_found <- tibble(date, life_stage, count) |> 
        filter(count > 0)
    
    encounters <- length(unique(vis_found$date))
    stages <- length(unique(vis_found$life_stage))
    totals <- sum(vis_found$count)
    
    if (encounters > 1 | stages > 1 | totals >= 6)
        return("established")
    if (totals > 0)
        return("suspected")
    return("absent/not detected")
}

pal <- met.brewer("Hokusai1", type="discrete")

parks_occ <- parks |> 
    st_drop_geometry() |> 
    group_by(genus, species, month) |> 
    # TODO: change to established=>=6 or >1 ls or >1 year (found in that month)
    summarise(occ=ifelse(sum(count > 0), "present", "absent")) |> 
    mutate(n=sum(occ == "present")) |> 
    ungroup() |> 
    unite("species", genus, species, sep=" ") |> 
    arrange(n)

p1 <- ggplot(parks_occ, aes(month(month, label=TRUE), fct_inorder(species), fill=occ)) +
    geom_tile(col="white", size=1.5) +
    scale_fill_manual(values=c(present=pal[2], absent=pal[7])) +
    guides(x=guide_axis(angle=45)) +
    theme_bw() +
    labs(subtitle=expression(paste(italic("When"), " were ticks found?")), x="Month", y=NULL, fill=NULL)

parks_occ <- parks |> 
    group_by(site, genus, species) |> 
    summarise(occ=occ_status_space(date, life_stage, count), .groups="drop") |> 
    unite("species", genus, species, sep=" ")

sc_counties <- st_read("geo-files/south-carolina-county-boundaries.shp") |> 
    st_transform(st_crs(parks))

p2 <- parks_occ |> 
    group_by(species) |> 
    mutate(n=sum(occ == "established"), species=paste0(species, " (", round(100 * n / 30, 1), "%)")) |> 
    ungroup() |> 
    arrange(desc(n)) |> 
    ggplot() +
    geom_sf(data=sc_counties, col="gray40") +
    geom_sf(aes(col=fct_relevel(occ, "established", after=Inf)), size=1.7) +
    facet_wrap(~fct_inorder(species)) +
    scale_color_manual(values=c(pal[7], pal[5], pal[2])) +
    guides(x=guide_axis(angle=45)) +
    labs(subtitle=expression(paste(italic("Where"), " were ticks found?")), col="Occupancy status") +
    theme_bw()

plot_grid(p1, p2, align="v", axis="l", ncol=1, rel_heights=c(1, 2))

ggsave("figs/figure1.pdf", width=9, height=6.5)
```

### Figure: county distribution of tick species, from shelter data

```{r}
shelters <- read_csv("data-proc/shelters-tidy-20-21.csv") |> 
    left_join(rename(sc_counties, county=COUNTYNM), by="county") |> 
    st_as_sf()
```

```{r}
shelters_occ <- shelters |> 
    filter(genus != "Unknown" & species != "unknown" & species != "unkown") |> 
    group_by(county, genus, species) |> 
    summarise(occ=occ_status_space(1, life_stage, count), .groups="drop") |> 
    unite("species", genus, species, sep=" ")

ggplot(shelters_occ) +
    geom_sf(data=sc_counties, col="gray40", fill=pal[7]) +
    geom_sf(aes(fill=fct_relevel(occ, "established", after=Inf))) +
    facet_wrap(~fct_inorder(species)) +
    scale_fill_manual(values=c(pal[5], pal[2])) +
    # guides(x=guide_axis(angle=45)) +
    labs(subtitle="County status from shelter collections", fill="Occupancy status") +
    theme_bw() +
    theme(legend.position="bottom")
```



### Table: raw counts of species X year X life_stage

```{r}
t1_dat <- parks |> 
    mutate(year=format(date, "%Y")) |> 
    group_by(year, genus, life_stage) |> 
    summarize(count=sum(count)) |>
    ungroup() |> 
    complete(year, genus, life_stage, fill=list(count=0)) |> 
    pivot_wider(names_from=life_stage, values_from=count)

colnames(t1_dat) <- c("", "", "Adult", "Larva", "Nymph")

kbl(t1_dat, booktabs=TRUE) |> 
    collapse_rows(1) |>
    add_header_above(c("Year"=1, "Genus"=1, "Life Stage"=3)) |> 
    kable_styling(latex_options="striped", stripe_color="blue") |> 
    kable_paper()
```

### Examine evidence for a random intercept/slope model 

### Relationship between monthly tick counts for different genera and life stages

Dermacentor has been removed from all of these plots since very few were collected.

#### Correlation of monthly counts between genus only

```{r, out.width="75%"}
parks |> 
    group_by(site, month, genus) |> 
    summarise(count=sum(count)) |> 
    ungroup() |> 
    filter(genus != "Dermacentor") |> 
    pivot_wider(names_from=genus, values_from=count, values_fill=0) |> 
    filter(Amblyomma < 500) |> 
    ggplot(aes(Amblyomma, Ixodes)) +
    geom_point(alpha=0.5) +
    geom_line(stat="smooth", method="lm", se=FALSE, col="forestgreen", alpha=0.5)
```

#### Lagged correlation of monthly counts between genus only

For each pair of genera $g$ and $h$, plot a regression of monthly counts $y_{gm}$ and $y_{hm-1}$.

```{r}
parks_mon <- parks |> 
    group_by(month, genus) |> 
    summarise(count=sum(count)) |> 
    ungroup() |> 
    filter(genus != "Dermacentor") |> 
    complete(month, genus, fill=list(count=0))

pmon_pairs <- full_join(parks_mon, parks_mon, by=character(), suffix=c("", "_prev")) |> 
    filter(as.numeric(month_prev) == as.numeric(month) - 1) |> 
    unite("label", genus_prev, genus, sep="_(t-1) x ") |> 
    mutate(label=str_c(label, "_t"))

ggplot(pmon_pairs, aes(count_prev, count)) +
    geom_point(alpha=0.5) +
    geom_line(stat="smooth", method="lm", se=FALSE, col="forestgreen", alpha=0.5) +
    facet_wrap(~label, scales="free", ncol=2) +
    labs(x="Count (previous month)", y="Count (current month)")
```

#### Now, separate by genus and life stage

Additionally filter out the nymphs and larvae for Ixodes since only 0, 5 in total.

```{r}
parks_mon <- parks |> 
    group_by(month, genus, life_stage) |> 
    summarise(count=sum(count)) |> 
    ungroup() |> 
    filter(genus != "Dermacentor") |> 
    complete(month, genus, fill=list(count=0)) |> 
    filter(genus != "Ixodes" | life_stage == "adult")
```

Plot pairs within the current month. Note that facet plots on opposing sides of the diagonal are equivalent.

```{r}
pmon_pairs <- full_join(parks_mon, parks_mon, by=character(), suffix=c("", "_prev")) |> 
    filter(
        as.numeric(month_prev) == as.numeric(month),
        (genus != genus_prev) | (life_stage != life_stage_prev) # don't compare to self since corr=1
    ) |> 
    unite("group", genus, life_stage, remove=TRUE) |> 
    unite("group_prev", genus_prev, life_stage_prev, remove=TRUE)

ggplot(pmon_pairs, aes(count_prev, count)) +
    geom_point(alpha=0.5) +
    geom_line(stat="smooth", method="lm", se=FALSE, col="forestgreen", alpha=0.5) +
    # facet_wrap(group_prev ~ group, scales="free", ncol=3, labeller=label_both) +
    facet_wrap(group_prev ~ group, scales="free", ncol=3) +
    labs(x="Count (current month)", y="Count (current month)")
```

Plot pairs of a monthly lag, including with previous month's counts from the same `genus x life_stage` group:

```{r, fig.width=7, fig.height=6}
pmon_pairs <- full_join(parks_mon, parks_mon, by=character(), suffix=c("", "_prev")) |> 
    filter(month_prev == month - months(1)) |> 
    unite("group", genus, life_stage, remove=TRUE) |> 
    unite("group_prev", genus_prev, life_stage_prev, remove=TRUE)

ggplot(pmon_pairs, aes(count_prev, count)) +
    geom_point(alpha=0.5) +
    geom_line(stat="smooth", method="lm", se=FALSE, col="forestgreen", alpha=0.5) +
    facet_wrap(group_prev ~ group, scales="free", ncol=4, labeller=label_both) +
    labs(x="Count (previous month)", y="Count (current month)")
```

We can also view this as a correlation matrix

```{r}
pivot_wider(pmon_pairs, names_from=group, values_from=count, values_fill=0) |> 
    pivot_wider(names_from=group_prev, values_from=count_prev, names_prefix="prev_", values_fill=0)
```

parks_mod |> 
    pivot_wider(c(month, site), names_from=life_stage, values_from=count, values_fn=sum) |> # agg. sp. etc
    replace_na(list(adult=0, nymph=0, larva=0)) |> 
    filter(adult < 50, nymph < 75, larva < 75) |>
    ggpairs(
        mapping=ggplot2::aes(col=month), 
        columns=c("adult", "nymph", "larva"), 
        lower=list(continuous="points"), 
        upper="blank"
    )

parks_mod |> 
    # select(-sex) |> 
    filter(life_stage == "adult", count < 100) |> 
    pivot_wider(c(month, site), names_from=genus, values_from=count, values_fn=sum) |> # agg. sp. from same genus
    select(-Dermacentor) |> 
    replace_na(list(Ixodes=0, Amblyomma=0, Dermacentor=0)) |> 
    ggpairs(
        mapping=ggplot2::aes(col=month), 
        columns=c("Ixodes", "Amblyomma"), 
        lower=list(continuous="points"), 
        upper="blank"
    )

ggplot(parks_mod, aes(mean_temp, log1p(count), col=genus)) +
    geom_point(alpha=0.6) +
    geom_smooth(method="lm", se=FALSE) +
    facet_wrap(~life_stage)

parks_mod |> 
    group_by(date, site) |> 
    summarize(num_ixodes=sum(count[genus=="Ixodes"]), num_amblyomma=sum(count[genus=="Amblyomma"])) |> 
    ggplot(aes(num_amblyomma, num_ixodes, col=format(date, "%Y"))) +
    geom_point()